A pipeline to analyse time-course gene expression data

### 2020-10-07 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

\hypertarget{looking-at-specific-clusters-in-details.}{%
\subsubsection{Looking at specific clusters in details.}\label{looking-at-specific-clusters-in-details.}}

@@

Most methods method to find the number of clusters with stability measures
only provide visual aids to guide the user. The first element often visualized
is the consensus matrix: the consensus matrix is an \(n \times n\) matrix that
stores the proportion of clustering in which two items are clustered together.
A perfect consensus matrix ordered such as all elements that belong to the
same cluster are adjacent to one another which show blocks along the diagonal
close to 1.

@@

Each column corresponds to a bootstrap sample, each row to a gene, and each
entry to the label found for that particular clustering. Thus, for each gene,
we have an assignment to a cluster over the 20 resampling runs. For example,
for \(k=5\):

\begin{longtable}[]{@{}ccccccccccccc@{}}

@@

\begin{minipage}[b]{0.13\columnwidth}\centering

@@

\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering

@@

@@

B11\strut
B12\strut
\begin{minipage}[b]{0.13\columnwidth}\centering
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B11\strut
B12\strut

@@

\begin{minipage}[t]{0.13\columnwidth}\centering

@@

\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\begin{minipage}[t]{0.13\columnwidth}\centering

@@

\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering

@@

\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\begin{minipage}[t]{0.13\columnwidth}\centering

@@

\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering

@@

\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\begin{minipage}[t]{0.13\columnwidth}\centering

@@

\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering

\begin{longtable}[]{@{}cccccccccccccc@{}}
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B13\strut
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B14\strut
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B15\strut
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B16\strut
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B17\strut
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B18\strut
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B19\strut
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B20\strut
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B21\strut
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B22\strut
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B23\strut
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B24\strut
\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
B25\strut
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering

@@

From this plot (Figure \ref{fig:plotModelExplorer}), we can deduce that \(k=5\)
is more stable than \(k=3\) and \(k=4\), but not as stable as \(k=2\). The model
explorer strategy, in addition to visualizing the diversity of the centroids,
can thus help assessing an adequate number of clusters.

Now, replot the same model explorer, but only for the clustering experiments
\(k=6\), \(k=7\), \(k=8\), \(k=9\) and \(k=10\) so that we can see more clearly the stability
measures in that range (Figure \ref{fig:plotModelExplorerZoom}).

@@

Let us first tackle the case of pathway enrichment analysis. We will leverage
the packages \texttt{biomaRt} \citep{durinck:biomart} and \texttt{KEGGprofile}
\citep{zhao:keggprofile} for this step. \texttt{KEGGprofile} is a package that facilitates
enrichment analysis on a set of genes labeled with the Ensembl annotation
\citep{ensembl2020} based on the set of biological pathways described in the KEGG
database\citep{KEGG}.

@@

##  [3] ggfortify_0.4.10            timecoursedata_0.99.17     
##  [5] moanin_0.99.0-902           SummarizedExperiment_1.18.2

>>>>

\hypertarget{looking-at-specific-clusters-in-detail.}{%
\subsubsection{Looking at specific clusters in detail.}\label{looking-at-specific-clusters-in-detail.}}

@@

Most methods to find the number of clusters with stability measures
only provide visual aids to guide the user. The first element often visualized
is the consensus matrix: the consensus matrix is an \(n \times n\) matrix that
stores the proportion of clustering in which two items are clustered together.
A perfect consensus matrix ordered such as all elements that belong to the
same cluster are adjacent to one another which show blocks along the diagonal
close to 1.

@@

Each column corresponds to a bootstrap sample, each row to a gene, and each
entry to the label found for that particular clustering. Thus, for each gene,
we have an assignment to a cluster over the 25 resampling runs. For example,
for the first 10 bootstrap samples and \(k=5\):

\begin{longtable}[]{@{}ccccccccccc@{}}

@@

\begin{minipage}[b]{0.15\columnwidth}\centering

@@

@@

\end{minipage} & \begin{minipage}[b]{0.06\columnwidth}\centering

@@

@@

\begin{minipage}[t]{0.15\columnwidth}\centering

@@

\end{minipage} & \begin{minipage}[t]{0.06\columnwidth}\centering
\begin{minipage}[t]{0.15\columnwidth}\centering

@@

@@

\end{minipage} & \begin{minipage}[t]{0.06\columnwidth}\centering
\begin{minipage}[t]{0.15\columnwidth}\centering

@@

@@

\end{minipage} & \begin{minipage}[t]{0.06\columnwidth}\centering
\begin{minipage}[t]{0.15\columnwidth}\centering

@@

\end{minipage} & \begin{minipage}[t]{0.06\columnwidth}\centering

@@

From this plot (Figure \ref{fig:plotModelExplorer}), we can deduce that \(k=5\)
is more stable than \(k=3\) and \(k=4\), but not as stable as \(k=2\). The model
explorer strategy, in addition to visualizing the diversity of the centroids,
can thus help assess an adequate number of clusters.

Now, replot the same model explorer, but only for the clustering experiments
\(k=6\), \(k=7\), \(k=8\), \(k=9\), and \(k=10\) so that we can see more clearly the stability
measures in that range (Figure \ref{fig:plotModelExplorerZoom}).

@@

Let us first tackle the case of pathway enrichment analysis. We will leverage
the packages \texttt{biomaRt} \citep{durinck:biomart} and \texttt{KEGGprofile}
\citep{zhao:keggprofile} for this step. \texttt{KEGGprofile} is a package that facilitates
enrichment analysis on a set of genes labeled with the Ensembl annotation
\citep{ensembl2020} based on the set of biological pathways described in the KEGG
database \citep{KEGG}.

@@

##  [3] ggfortify_0.4.10            timecoursedata_0.99.18     
##  [5] moanin_0.99.8               SummarizedExperiment_1.18.2
